---
title: R Graphics
Author: NG
Date: today
Output: html_document
---

# base, grid, lattice, ggplot2

# Scatter Plots

## base
* plot() function
* col, pch parameters to adjust colour and style
```{r}
plot(iris$Petal.Length, iris$Petal.Width, 
     col = "red", pch = 2)
```

* Use layout() function to display multiple graphics
```{r}
par(mar = c(3, 3, 0.5, 0.5), oma = rep.int(0, 4), mgp = c(2, 1, 0))
species = levels(iris$Species)
plot_numbers = seq_along(species)
layout(matrix(plot_numbers, ncol = 3, byrow = TRUE))
for(s in species) {
  data.by.species = subset(iris, Species == s)
  with(data.by.species, plot(Sepal.Length, Sepal.Width))
}
```

## lattice
* xyplot()

## ggplot2
* ggplot() function with parameter geom_plot()

# Line Plots

## base
* plot() function with parameter type = "1"
* lines() function to add additional line
```{r}
set.seed(1)
y1 <- cumsum(rnorm(100))
x <- seq_along(y1)
plot(x, y1, type = "l", ylim = c(-5, 20))
y2 <- y1 + abs(rnorm(100, 3))
lines(x, y2, col = "blue")
```

## lattice
* xyplot() function with parameter type = "1"

## ggplot2
* ggplot() function with parameter geom_line()

# Histograms

## base
* hist() function
* Add parameter freq = FALSE if want to have total area = 1
* Add parameter breaks to define number of bar
```{r}
hist(iris$Sepal.Width)
hist(iris$Sepal.Width, breaks = c(2, 2.7, 3, 3.5, 4.5))
```

## lattice
* histogram() function
* y axis adjust by parameter type
* 3 types can be choose: percent, count, density
* parameter break is the same as hist() function

## ggplot2
* ggplot() function with parameter geom_histogram()
* Parameter bins in geom_histogram() can define number of bin

# Box Plots

## base
* boxplot() function
```{r}
boxplot(InsectSprays$count)
# draw for count according to spray
boxplot(count ~ spray, data = InsectSprays)
# define data frame to order
my.InsectSprays <- within(InsectSprays,  spray <- (reorder(spray, count, median)))
boxplot(count ~ spray, data = my.InsectSprays)
```

## lattice
* bwplot() function
* Similar to boxplot() function

## ggplot2
* ggplot() function with parameter geom_boxplot()

# Bar Plots

## base
* barplot() function
* Use parameter names.arg to define the name of each bar
* Use parameter horiz = TRUE to switch to horizontal bar
* Word orientation have to apply par(las = 1, mar = c(3, 5, 1, 1))
* las = 1 means horizontal
* mar means margin from below, left, top to right
```{r}
gear.table <- table(mtcars$gear)
barplot(gear.table, names.arg = c("Three", "Four", "Five"))
par(las = 1, mar = c(3, 5, 1, 1))
barplot(gear.table, names.arg = c("Three", "Four", "Five"), horiz = TRUE)
```

* Combine to data set in one chart
* Add parameter beside = TRUE to seperate
```{r}
gear.table2 <- table(mtcars$vs, mtcars$gear)
barplot(gear.table2)
barplot(gear.table2, beside = TRUE)
```

## lattice
* barchart() function
* Add parameter stack = FALSE to seperate bars

## ggplot2
* ggplot() function with parameter geom_bar()
* Add parameter position = "dodge" in geom_bar() to split bars
* Add parameter coord_flip() to create horizontal bar

# Pie chart

## base
* pie() function
* Use par(mar = c(1 ,1 ,1 ,1)) to control size
* Add parameter labels to define names
```{r}
pie(gear.table, labels = c("Three", "Four", "Five"))
```

## lattice
* No Function Provided

## ggplot2
* Create by combining bar chat and then cut to circle
* Use coord_polar to transform
* Add theme_void to hide axes of bar chart
* Syntax as below
```{r, eval = FALSE}
pie <- ggplot(mtcars,
  aes(x = factor(1), fill = factor(gear))) +
  geom_bar(width = 1)
pie # create bar chart
pie <- pie + coord_polar(theta = "y")
pie # transform to circular shape
pie <- pie + theme_void()
pie # hide axes
```

# Other plotting packages

* vcd: 
* plotrix: 
* hexbin: 
* scatterplot3d: 
* misc3d: 
* rgl: Use OpenGL to plot 3D graph 
* rggob: Use GGobi in R
* igraph: 
* plotly: Plotly for R 

* Check plotting packages from CRAN Task View