---
title: R Packages
Author: NG
Date: today
output: html_document
---

# Basic

* Use install.packages() function to install packages
* Install "devtools" if want to install packages from GitHub
* Then use install_github() function to install

* Use update.packages() to update packages
* Add arguement ask = FALSE to omit notice
* Can add packages name as arguement to update specific packages only

* Use library() function to import package
* Use require() function to import package
* library() will cause error if specified package does not exit
* require() will return TRUE or FALSE but not error

* Use search() function to list loaded package
```{r}
search()
```

* Use installed.packages() function to list installed packages
* Return will be data frame
* Use view() function to display in table
* Use R.home("library") function or .Library variable to check the directory
* Use path.expand("~") and Sys.getenv("HOME") to check the path
```{r}
.Library
path.expand("~")
Sys.getenv("HOME")
```

* .libPaths() function to list the path when searching packages
* Adjust path to search in other path
* Can add additional path to R_LIBS
```{r}
# path <- .libPaths() ; .libPaths(c(path, "D:/my_lib"))
.libPaths()
```

# Suggested packages

* lattice: provided useful function for plotting
* ggplot2: provided powerful function for plotting




