---
title: Sample R Project
subtitle: from G.T. Wang blog
Author: NG
Date: today
output: html_document
---

# Project: 2015 年臺南市本土登革熱疫情狀況

## 1. Import data
Data name: dengue-20151107-big5.csv

```{r}
# To show the working directory
getwd()
# Import data
dengue = read.csv("dengue-20151107-big5.csv")
```

## 2. Check data
```{r}
str(dengue)
summary(dengue)
```

## 3. Filter data
```{r}
filter.idx1 <- dengue$緯度座標 > 22.8 & dengue$緯度座標 < 23.5
filter.idx2 <- dengue$經度座標 > 120 & dengue$經度座標 < 120.6
dengue.tn <- dengue[filter.idx1 & filter.idx2, ]
str(dengue.tn)
summary(dengue.tn)
```

## 4. There are some repeated levels due to space
```{r}
levels(dengue.tn$區別)
```

## 5. To correct the data
```{r}
dengue.tn[dengue.tn$區別 == "北　區", ]$區別 <- "北區"
dengue.tn[dengue.tn$區別 == "東　區", ]$區別 <- "東區"
dengue.tn[dengue.tn$區別 == "南　區" | dengue.tn$區別 == "南    區", ]$區別 <- "南區"
dengue.tn[dengue.tn$區別 == "永康區 ", ]$區別 <- "永康區"
```

## 6. Rebuild factor to remove unwanted levels
```{r}
dengue.tn$區別 <- factor(dengue.tn$區別)
levels(dengue.tn$區別)
```

## 7. Start to analyse
```{r}
hist(as.Date(dengue.tn$確診日), breaks = "weeks",
  freq = TRUE, main = "登革熱每週病例數", xlab = "日期",
  ylab = "病例數", format = "%m/%d")
```

* To view monthly data
```{r}
dengue.tn$month <- format(as.Date(dengue.tn$確診日), "%m")
table(dengue.tn$month)
barplot(table(dengue.tn$month), xlab = "月份", ylab = "病例數",
  main = "登革熱每月病例數")
```

### Plot by ggplot2
```{r}
library(ggplot2)
library(scales)
ggplot(dengue.tn, aes(x=as.Date(確診日))) +
  stat_bin(binwidth=7, position="identity") +
  scale_x_date(breaks=date_breaks(width="1 month")) +
  theme(axis.text.x = element_text(angle=90)) +
  xlab("日期") + ylab("病例數") +
  ggtitle("登革熱每週病例數")
```

## 8. To view by districts
```{r}
dengue.region.summary <- sort(summary(dengue.tn$區別), decreasing = FALSE)
dengue.region.summary
```

## 9. Plot
```{r}
barplot(dengue.region.summary, las = 2, horiz = TRUE,
  main = "各行政區病例統計", xlab = "病例數")
```

* By pie chart
```{r}
pie(dengue.region.summary)
```

## 10. Sort the 5 most series districts
```{r}
dengue.top.reg <- dengue.tn[
  dengue.tn$區別 == "北區" |
  dengue.tn$區別 == "中西區" |
  dengue.tn$區別 == "南區" |
  dengue.tn$區別 == "東區" |
  dengue.tn$區別 == "永康區", ]
```

## 11. Plot the 5 districts data
```{r}
ggplot(dengue.top.reg, aes(x=as.Date(確診日))) +
  stat_bin(binwidth=7, position="identity") +
  scale_x_date(breaks=date_breaks(width="1 month")) +
  theme(axis.text.x = element_text(angle=90)) +
  xlab("日期") + ylab("病例數") +
  ggtitle("登革熱每週病例數") + facet_grid(區別 ~ .)
```

* Plot monthly
```{r}
ggplot(dengue.top.reg, aes(x=as.Date(確診日))) +
  stat_bin(breaks=as.numeric(seq(as.Date('2015-1-1'),
    as.Date('2015-12-1'), '1 month')), position="identity") +
  scale_x_date(breaks=date_breaks(width="1 month")) +
  theme(axis.text.x = element_text(angle=90)) +
  xlab("日期") + ylab("病例數") +
  ggtitle("登革熱每月病例數") + facet_grid(區別 ~ .)
```

# More
* If someone live in (22.997088, 120.201771), sort data within 400m
* Function to calculate distant of two point on Earth
```{r}
earthDist <- function (lon1, lat1, lon2, lat2)
{
  rad <- pi/180
  a1 <- lat1 * rad
  a2 <- lon1 * rad
  b1 <- lat2 * rad
  b2 <- lon2 * rad
  dlon <- b2 - a2
  dlat <- b1 - a1
  a <- (sin(dlat/2))^2 + cos(a1) * cos(b1) * (sin(dlon/2))^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  R <- 6378.145
  d <- R * c
  return(d)
}
```

* To sort data
```{r}
home.pos <- c(22.997088, 120.201771) # (緯度, 經度)
home.dist <- earthDist(dengue.tn$經度座標, dengue.tn$緯度座標, home.pos[2],  home.pos[1])
home.idx <- home.dist <= 0.4;
dengue.home <- dengue.tn[home.idx, ]
```

* View monthly data
```{r}
table(dengue.home$month)
```

* Plot
```{r}
barplot(table(dengue.home$month), xlab = "月份", ylab = "病例數",
  main = "登革熱每月病例數（特定區域）")
```

# There are some sections using ggmap to plot data on map, but ggmap require api, so I just omit those section.